Adaptive Sequential Model Selection

### 2015-05-04 paper.tex
fixed some bugs in paper

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\hK}{\widehat{K}}
\newcommand{\hk}{\hat{k}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}

@@

\title{Adaptive Sequential Model Selection}
\author{William Fithian, Jonathan Taylor, Rob Tibshirani, \and Ryan Tibshirani}
\maketitle

@@

\begin{example}[Bivariate Regression]
  Consider forward stepwise selection in a regression model with $n=p=2$, with identity design $X = I_2=\begin{pmatrix} 1 & 0 \\ 0 & 1\end{pmatrix}$ and known $\sigma^2=1$. We perform one step of forward stepwise and then test the null model with no variables against the model with one variable.

We will choose variable 1 first on the selection event $A=\{|Y_1| > |Y_2|\}$, which is shown in yellow in Figure~\ref{fig:condSets}. In that case,
\[
M_0:\; Y\sim \cN(0,I_2), \quad\text{ and } 
M_1:\; Y \sim \cN\left(\frac{\mu_1}{0}, \; I_2\right).
\]
The selected-model test compares $Y_1$ to its distribution under $M_0$ conditional on $A$, a test of $H_0:\;\mu_1=0$ in model $M_1$.

@@

Figure~\ref{fig:condSets} shows the conditioning sets for each model when $Y=(2.9,2.5)$. Next to it, Figure~\ref{fig:nullDists} shows the null distribution for the test statistic $Y_1$ in each case. The $p$-values for the selected and saturated models are 0.007 and 0.3, respectively.
\end{example}

The case illustrated in Figure~\ref{fig:condSets} exhibits a phenomenon that has been remarked upon elsewhere in the literature of saturated-model tests: when there are near-ties between strong variables that are competing to enter the model, the winning variable may have a very weak $p$-value. \WFcomment{add references}. Figure~\ref{fig:}

While the selected-model $p$-value is not always smaller than the saturated-model $p$-value, the selected-model test is strictly more powerful when the selected model is correct; i.e., when $\mu_2=0$. Figure~\ref{fig:powCurves_0} shows the power curve for each test when $\mu_2=0$. While the selected-model test is more powerful, the difference between the two is relatively mild. The difference is much more pronounced when $\mu_2=4$, as shown in Figure~\ref{fig:powCurves_4}.

@@

\subsection{The Excluded-Variable Irrelevance Property}

We say that a selection algorithm $M_{[d]}(\cdot)$ has the {\em excluded-parameter irrelevance condition} (henceforth EPIC) if, given $M_k(Y)=m_k$, the sequence of models $M_{[k]}$ depends only on sufficient statistics of $m_k$. That is

Ever-active coefficients in a regularization path are 
\begin{proposition}
  Let $M_\infty$ be a model with parameters $\theta\in\Theta\sub \R^p$, and likelihood $L(\theta; Y)$. For $k\in [d]$, let $P_k(\theta)$ denote a regularizing penalty to use at step $k$. Define
%  \begin{align*}
%    \hat\theta^{k}(Y) &\triangleq 
%    \argmin_{\theta\in\Theta} -\log L(\theta; Y) + P_k(\theta) \\
%    E_k(Y) &\triangleq \left\{j:\; \hat\theta_j^\ell \neq 0 
%      \text{ for any \ell \leq k} \right\}\\ 
%    M_k(Y) &\triangleq \{F\in M_\infty:\; 
%    \theta_j = 0,\; \forall j \notin E_k\}
%  \end{align*} 
  Then $M_d(\cdot)$ satisfies EPIC.
\end{proposition}
\WFcomment{Need to be able to prune away the times when $M_k=M_{k-1}$; otherwise how do you construct $p_k$ for that time?}

@@

\begin{proposition}

\end{proposition}

Let $\sF_{k,\ell}$ denote the $\sigma$-algebra generated by $M_{[k]}$ and $p_{[\ell]}$.
\begin{align*}
  \sF_{k,\ell} &= \sF(M_{[k]},p_{[\ell]})\\
  \sF_0 &\underlabel_{\text{selection } 1} \sF_{1,0} \underlabel_{\text{inference } 1}
  \sF_{1,1} \;\;\sub \cdots \sub\;\;
  \sF_{d-1,d-1} \underlabel_{\text{selection } d} \sF_{d,d-1}
  \underlabel_{\text{inference } d} \sF_{d,d}
\end{align*}

@@

>>>>

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\hK}{\widehat{K}}
\newcommand{\hk}{\hat{k}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}

@@

\title{Adaptive Sequential Model Selection}
\author{William Fithian, Jonathan Taylor, Rob Tibshirani, and Ryan Tibshirani}
\maketitle

@@

\begin{example}[Bivariate Regression]\label{ex:bivariate}
  Consider forward stepwise selection in a regression model with $n=p=2$, with identity design $X = I_2=\begin{pmatrix} 1 & 0 \\ 0 & 1\end{pmatrix}$ and known $\sigma^2=1$. We perform one step of forward stepwise and then test the null model with no variables against the model with one variable.

We will choose variable 1 first on the selection event $A=\{|Y_1| > |Y_2|\}$, which is shown in yellow in Figure~\ref{fig:bv_condSets}. In that case,
\[
M_0:\; Y\sim \cN(0,I_2), \quad\text{ and } 
M_1:\; Y \sim \cN\left(\frac{\mu_1}{0}, \; I_2\right).
\]
The selected-model test compares $Y_1$ to its distribution under $M_0$ conditional on $A$, a test of $H_0:\;\mu_1=0$ in model $M_1$.

@@

Figure~\ref{fig:bv_condSets} shows the conditioning sets for each model when $Y=(2.9,2.5)$. Next to it, Figure~\ref{fig:bv_nullDists} shows the null distribution for the test statistic $Y_1$ in each case. The $p$-values for the selected and saturated models are 0.007 and 0.3, respectively. These two plots are reproduced from \citet{fithian2014optimal}, in which the same example was presented in less detail.
\end{example}

\begin{figure}
  \centering
  \begin{subfigure}[t]{.4\textwidth}
    % source code: bivariateSelVSat.R
    \includegraphics[width=\textwidth]{figs/bivariateSelVSat_condSets.pdf}
    \caption{\WFcomment{Copied directly from FST14; reword} 
      For $Y=(2.9,2.5)$, the selected-model conditioning set is
      $A=\{y:\;|y_1|>|y_2|\}$, a union of quadrants,
      plotted in yellow. The saturated-model conditioning set
      is ${\{y:\; y_2=2.5\}\cap A} = {\{y:\;y_2=2.5, |y_1|>2.5\}}$,
      a union of rays, plotted in brown.}
    \label{fig:bv_condSets}
  \end{subfigure}
  \hspace{.1\textwidth}
  \begin{subfigure}[t]{.4\textwidth}
    % source code: bivariateSelVSat.R
    \includegraphics[width=\textwidth]{figs/bivariateSelVSat_nullDists.pdf}
    \caption{\WFcomment{Copied directly from FST14; reword} 
      Conditional distributions of $Y_1$ under
      $H_0:\mu_1 = 0$. Under the hypothesis
      $\mu=0$, the realized  $|Y_1|$ is  quite large given $A$,
      giving  $p$-value 0.015. By contrast, $|Y_1|$ is not too large
      given $A \cap \{y:\; y_2=Y_2\}$, giving
      $p$-value 0.3.}
  \end{subfigure}
  \caption{\WFcomment{Copied directly from FST14; reword} 
    Contrast between the saturated-model and selected-model tests
    in Example~\ref{ex:bivariate}, in which we fit a one-sparse model with
    design matrix $\bX=I_2$. The selected-model test is based
    on  $\L_0(Y_1 \gv A)$, whereas the saturated-model test is based
    on $\L_0(Y_1  \gv  Y_2, A)$.}
  \label{fig:bv_nullDists}
\end{figure}

The case illustrated in Figure~\ref{fig:bv_condSets} exhibits a phenomenon that has been remarked upon elsewhere in the literature of saturated-model tests: when there are near-ties between strong variables that are competing to enter the model, the winning variable may have a very weak $p$-value. \WFcomment{add references}. Figure~\ref{fig:bv_rocCurve} displays the cumulative distribution function for the first $p$-value when $\mu=\binom{4}{4}$, a very strong signal. While the selected model test has near perfect power, it is not uncommon for the saturated model test to produce large $p$-values, even in the range of 0.5-0.9. These large $p$-values arise exactly when there is a near tie between the variables.

\begin{figure}
  \centering
  % source code: bivariateSelVSat.R
  \includegraphics[width=.5\textwidth]{figs/bivariateSelVSat_rocCurve.pdf}
  \caption{\WFcomment{Write caption here.}}
  \label{fig:bv_rocCurve}
\end{figure}

Results in \citet{fithian2014optimal} show that the selected-model test is strictly more powerful when the selected model is correct; i.e., when $\mu_2=0$. Figure~\ref{fig:bv_powCurves_0} shows the power curve for each test when $\mu_2=0$. While the selected-model test is more powerful, the difference between the two is relatively mild. Interestingly, the difference is much more pronounced when $\mu_2=4$, as shown in Figure~\ref{fig:bv_powCurves_4}.

\begin{figure}
  \centering
  \begin{subfigure}[t]{.4\textwidth}
    % source code: bivariateSelVSat.R
    \includegraphics[width=\textwidth]{figs/bivariateSelVSat_powCurves_0.pdf}
    \caption{\WFcomment{Write caption here.}}
    \label{fig:bv_powCurves_0}
  \end{subfigure}
  \hspace{.1\textwidth}
  \begin{subfigure}[t]{.4\textwidth}
    % source code: bivariateSelVSat.R
    \includegraphics[width=\textwidth]{figs/bivariateSelVSat_powCurves_4.pdf}
    \caption{\WFcomment{Write caption here.}}
  \end{subfigure}
  \caption{\WFcomment{Write caption here.}}
   \label{fig:bv_powCurves_4}
\end{figure}

@@

We say that a selection algorithm $M_{[d]}(\cdot)$ satisfies the {\em sub-path sufficiency principle} (henceforth SPSP) if, given $M_k(Y)=m_k$, the sequence of models $M_{[k]}$ depends only on sufficient statistics of $m_k$. That is

A model selection procedure derived from the ``ever-active'' coefficients of a regularized likelihood path always satisfies the SPSP. Let $M_\infty$ be a model with parameters $\theta\in\Theta\sub \R^p$, and likelihood $L(\theta; Y)$. For $r=0,\ldots,R$, let $P_r(\theta)$ denote a sequence of regularizing penalties, and define
\begin{align}\label{eq:regPathDef_start}
  \hat\theta^{r}(Y) &\triangleq 
  \argmin_{\theta\in\Theta} -\log L(\theta; Y) + P_r(\theta) \\
  \tE_r(Y) &\triangleq \left\{j:\; \hat\theta_j^s \neq 0 
    \text{ for any } s \leq r \right\}
\end{align}
Note that while the sets $\tE_r$ are nested by definition, we could have $\tE_r = \tE_{r+1}$ for most values of $r$. Let $r(0)=0$, and for $k>0$ define
\begin{equation}
r(k) = \min\{s:\; \tE_s \neq \tE_{r(k-1)}\}.
\end{equation}
That is, the $r(k)$ are the indices where the active set changes. Let $E_k = \tE_{r(k)}$, and take
\begin{equation}\label{eq:regPathDef_end}
M_k \triangleq \{F\in M_\infty:\; \theta_j = 0,\; \forall j \notin E_k\}
\end{equation}
Model paths described in this way satisfy the SPSP, as we see next.

\begin{proposition}[Regularized Likelihood Paths Satisfy SPSP]
  Model paths specified as in (\ref{eq:regPathDef_start}--\ref{eq:regPathDef_end}) satisfy the sub-path sufficiency principle.
\end{proposition}
\WFcomment{Actually, you just keep the times when the model changes... Need to recast notation a bit; indices of $\lambda$ (or $P$, here) need not match up with indices of model.}

@@

\begin{proposition}[Forward Stepwise Paths Satisfy SPSP]
  \WFcomment{Generalized forward stepwise}
\end{proposition}

\WFcomment{This is a filtration interpretation when you have the appropriate sufficiency properties.} Let $\sF_{k,\ell}$ denote the $\sigma$-algebra generated by $M_{[k]}$ and $p_{[\ell]}$.
\begin{align*}
  \sF_{k,\ell} &= \sF(M_{[k]},p_{[\ell]})\\
  \sF_0 &\underlabel_{\text{selection } 1} \sF_{1,0} \underlabel_{\text{inference } 1}
  \sF_{1,1} \;\;\sub \cdots \sub\;\;
  \sF_{d-1,d-1} \underlabel_{\text{selection } d} \sF_{d,d-1}
  \underlabel_{\text{inference } d} \sF_{d,d}
\end{align*}

\subsection{}

@@

\section{Simulation: Principal Components Analysis}

\WFcomment{Can we do this one?}
